---
title: "Project"
author: "Saurabh Singh"
date: "October 20, 2015"
output: html_document
---

%%%%%%%%%%%% A Binary Classification Model for predicting arrivals delays %%%%%%%%%%%%%%

Problem Statement: Build a binary classification model for predicting arrival delays.
Data: Use the US Dept. of Transportation on-time arrival data for non-stop domestic flights by major air carriers to predict arrival delays.
Data used here: Period = January 2015,Rows = 17802,Columns = 22.


1. Load the csv file in R
2. Remove all the blanks(NULL values) from the Arrival Delay column(response)
3. As we only want to know whether a fight will get delayed or not in future,in Arrival       Delay column,all values greater than 0 will be changed to 1(i.e.delay) and remain          0(i.e. not delay) otherwise.

```{r}

data <- read.csv("Flight.csv")

data1<- data[-which(is.na(data$ARR_DELAY_NEW)), ]

data1$ARR_DELAY_NEW[data1$ARR_DELAY_NEW > 0] <- 1

```

1. I will now find the best combination of predictors(variables) to make a model which can    predict the response variable(arrival delay) correctly.
2. I have considered only seven variables here which looks relevant to me and those           are-DAY_OF_MONTH,DAY_OF_WEEK,AIRLINE_ID,FL_NUM,ORIGIN_AIRPORT_ID,DEST_AIRPORT_ID,ARR_T     IME. 
3. regsubsets() function is used on training data set to perform best subset selection of     predictors.
4. (nvmax=7) maximum model size value used here is 7 as there are maximum seven predictors.
5. Then I calculated the validation set error for the best model of each model size.
6. For each model size, the coefficients of best model is used to predict the response        value and then the test MSE(mean square error) is calculated for each model.
7. At the end,I found out that the model with 6 variables is best as it gives the least       MSE.


```{r}

# Choosing among models using Cross Validation Technique  

# predict function to find the predicted value for each model size is created   

prediction <- function(object,newdata,id){
  form <- as.formula(object$call[[2]])
  mat <- model.matrix(form ,newdata)
  coefficient <- coef(object ,id=id)
  xvars <- names(coefficient)
  mat[,xvars ]%*% coefficient
  }

library(leaps)
k <- 10          # number of folds used in cross validation
set.seed(1)
folds <- sample(1:k,nrow(data1),replace=TRUE)
cv.errors <- matrix(NA,k,7,dimnames=list(NULL,paste(1:7)))

# Loop to find the best model for each model size is created
# Matrix of k cross 7 (or 10X7) is formed here
# In matrix (i,j)th elememt corresponds to the test MSE for the ith cross-validation fold  
# for the best j-variable model. 

for (j in 1:k){
  best.fit <- regsubsets(ARR_DELAY_NEW ~ DAY_OF_MONTH+DAY_OF_WEEK+AIRLINE_ID+FL_NUM+
                 ORIGIN_AIRPORT_ID+DEST_AIRPORT_ID+ARR_TIME,data=data1[folds!=j,],nvmax=7)

  for (i in 1:7){
    pred <- prediction(best.fit,data1[folds==j,],id=i)
    cv.errors[j,i] <- mean((data1$ARR_DELAY_NEW[folds==j]-pred)^2)
  }
}

mean.cv.errors <- apply(cv.errors,2,mean)                                                 

# apply() function to average over the columns of the matrix in order to obtain a vector 
# for which the jth element is the cross-validation error for the j-variable model 

mean.cv.errors

par(mfrow=c(1,1))
plot(mean.cv.errors,type='b')

reg.best=regsubsets(ARR_DELAY_NEW ~ DAY_OF_MONTH+DAY_OF_WEEK+AIRLINE_ID+FL_NUM+
                 ORIGIN_AIRPORT_ID+DEST_AIRPORT_ID+ARR_TIME,data=data1,nvmax=7)

coef(reg.best,6)  # predictors associated with response variable


```

1. Divide the whole data set in 80:20(or 4:1) ratio as training and validation(test) data     set.
2. Training set has 13160 rows and validation set has 3291 rows.
3. Fit a logistic regression model using only the training data set.
4. Predict the probabilities that a flight will arrive late in the validation data set.
5. Only 6 predictors are used in glm function as from the last part I got to know that only    six out of seven predictors are associated and would be helpful to predict response.
6. Probabilities above 0.5 are considered to be a delay.  


```{r}

tr <- sample(1:nrow(data1), size =13160 , replace=F)
train <- data1[tr, ]      # training data set
test <- data1[-tr, ]      # validation data set

glm.fit <- glm(ARR_DELAY_NEW ~ DAY_OF_MONTH+DAY_OF_WEEK+FL_NUM+
                 ORIGIN_AIRPORT_ID+DEST_AIRPORT_ID+ARR_TIME,data=train,                                   family = binomial)     # generalized linear model function
summary(glm.fit)


glm.probs <- predict(glm.fit,newdata=test,type="response")   # predict the probability

glm.pred <- rep(0,3291)
glm.pred[glm.probs > 0.5] <- 1      # consider as delay if probabilty is above 0.5a  

status.real <- test$ARR_DELAY_NEW

table(glm.pred,status.real)

mean(glm.pred==status.real)    # accuracy %

mean(glm.pred!=status.real)    # error %


```
Conclusion: 

1. Test error rate is 39.28 % and the accuracy rate is 60.71 % which means almost 61% of      the arrival delays could be predicted correctly.
2. We can improve the accuracy rate if we choose a better model,good predictors or by         changing the ratio of training and test data sets.

